---
title: 'An Introduction to INCA'
author:
  - name: Jingqi Duan
  - name: Sunduz Keles
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{INCA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r eval=TRUE, echo=FALSE}
suppressWarnings(library(knitr))
suppressMessages(library(data.table))
```

# Introduction

We have developed a multi-step genetic variant scoring approach that leverages the ENCODE RNA-binding protein (RBP) data from the [eCLIP-seq collection](https://www.encodeproject.org/search/?type=Experiment&control_type!=*&status=released&perturbed=false&assay_title=eCLIP&replicates.library.biosample.donor.organism.scientific_name=Homo+sapiens&biosample_ontology.term_name=K562&biosample_ontology.term_name=HepG2&assembly=hg19&files.file_type=bed+narrowPeak&files.file_type=bam&assay_slims=RNA+binding) together with [ClinVar](https://www.ncbi.nlm.nih.gov/clinvar/) and integrates multiple computational approaches to evaluates variant impacts on RBP activities. 

This document provides an introduction to the **INCA** package. **INCA** derives three scores for the effect of each variant on each RBP activity using available cell lines: (i) ClinVar-quantiled SeqWeaver scores; (ii) a score of allelic effect on RBP binding derived from a pre-computed library of RBP eCLIP-seq experiments; and (iii) a gene impact score for the gene that the SNV is proximal to based on a pre-computed library of accompanying RNA-seq experiments (wild type and RBP knockdown by shRNA).

# Installation

**INCA** will be submitted to Bioconductor. Currently, **INCA** can be downloaded and installed in R by: 

```{r eval=FALSE}
devtools::install_github('jduan607/INCA')
```

**INCA** depends on the following \R{} packages:

(a) `r CRANpkg("data.table")` is used for reading, writing, and formatting large data.

(b) `r CRANpkg("foreach")` is used for 

(c) `r CRANpkg("HGNChelper")` is used for 

Parallel backend is not required to register, but it will reduce computing time. The parallel backend can be registered in the following way:

```{r eval=FALSE}
library(doMC)
registerDoMC()
```

If not registered, set `parallel`=FALSE (as the default).

# Input

**INCA** takes as input a variant file with the following fields: _Chr_, _Pos_, _Start_, _End_, _Ref_, _Alt_, and _Gene_. 

Chr|Pos|Start|End|Ref|Alt|Gene
---|---|---|---|---|---|---
chr7|75607155|75607155|75607155|A|G|POR
chr7|75611756|75611756|75611756|C|T|POR
...||||||
chr7|75636240|75636240|75636240|T|C|STYXL1


## Whole-genome Sequencing (WGS) Data

We extracted the required information (Chr, Pos, Ref, Alt, Zyg) from the VCF files of WGS in HepG2 and K562 cell lines from the ENCODE project:
  + HepG2: ENCFF713BPG
  + K562: ENCFF752OAX
    
The corresponding code can be found in 'ExtractVCFInfo.py'.

Note: Zyg = 0 - Homozygous reference; 1 - Heterozygous; 2 - Unknown; 3 - Homozygous alternative

# Main Example in the Paper

Variants in the example are in high linkage-disequilibrium with the SNP _rs1057868_. INCAscore is computed using the data for the RBP _HNRNPK_.

## Load required data

```{r eval=FALSE}
library(INCA)

directory = 'https://raw.github.com/jduan607/INCA/master'

## GWAS
variants = fread(file.path(directory, 'GWAS', 'final_data.txt'))
SW = fread(file.path(directory,'GWAS','seqweaver_results.tsv'))
SW = SW[,c(1:8, grep('HepG2|K562',colnames(SW))),with=FALSE] # the example focuses on HepG2 and K562

## WGS
wgs1 = fread(file.path(directory,'ENCODE_WGS','HepG2_WGS.txt.gz'))
wgs2 = fread(file.path(directory,'ENCODE_WGS','K562_WGS.txt.gz'))

## DGE
dge1 = fread(file.path(directory,'ENCODE_shRNA/DGE','HNRNPK_HepG2_DGE.txt.gz'))
dge2 = fread(file.path(directory,'ENCODE_shRNA/DGE','HNRNPK_K562_DGE.txt.gz'))

## ENCODE - peak
peak1 = fread(file.path(directory,'ENCODE_eCLIP/PeakSignals','HNRNPK_HepG2_PeakSignals.txt.gz'))
peak2 = fread(file.path(directory,'ENCODE_eCLIP/PeakSignals','HNRNPK_K562_PeakSignals.txt.gz'))

## ENCODE - read counts (Load if needed)
exp1.1 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_HepG2_NormRC_Rep1.txt.gz')) 
exp1.2 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_HepG2_NormRC_Rep2.txt.gz')) 
ctrl1 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_HepG2_NormRC_Ctrl.txt.gz'))

rc1.1 = compareRCToControl(exp1.1, ctrl1)
rc1.2 = compareRCToControl(exp1.2, ctrl1)

exp2.1 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_K562_NormRC_Rep1.txt.gz')) 
exp2.2 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_K562_NormRC_Rep2.txt.gz')) 
ctrl2 = fread(file.path(directory,'ENCODE_eCLIP/NormRC','HNRNPK_K562_NormRC_Ctrl.txt.gz'))

rc2.1 = compareRCToControl(exp2.1, ctrl2)
rc2.2 = compareRCToControl(exp2.2, ctrl2)
```

## Align genotypes of K562 and HepG2 cell lines to variants

```{r eval=FALSE}
variants = alignCellGenoToVar(variants, wgs1, wgs2, c('HepG2','K562'))
```

## ClinVar-quantiled SeqWeaver scores

```{r eval=FALSE}
variants = scoreClinVarQSW(variants, SW, 'HNRNPK', empirical=1)
```

## eCLIP-seq allelic effects

```{r eval=FALSE}
epg1 = list(list(peaks=peak1[signalValue_IDR>0,]),
            list(peaks=peak1, threshold=0.5), # optional
            list(counts=rc1.1, peaks=peak1[signalValue_IDR>0,], threshold=c(0.1,0.8)), # optional
            list(counts=rc1.2, peaks=peak1[signalValue_IDR>0,], threshold=c(0.1,0.8))) # optional
            
epg2 = list(list(peaks=peak2[signalValue_IDR>0,]),
            list(peaks=peak2, threshold=0.5), # optional
            list(counts=rc2.1, peaks=peak2[signalValue_IDR>0,], threshold=c(0.1,0.8)), # optional
            list(counts=rc2.2, peaks=peak2[signalValue_IDR>0,], threshold=c(0.1,0.8))) # optional

# parallel = FALSE if no parallel backend registered
variants = scoreAllelicEffect(variants, epg1, epg2, c('HepG2','K562'), window=15, parallel=TRUE)
```

### RBP-SNV impact on gene expression

```{r eval=FALSE}
epg1 = list(dge1[,.(gene,q_value)], # first element must be DE
           list(peaks=peak1, threshold=1)) 
epg2 = list(dge2[,.(gene,q_value)], 
           list(peaks=peak2, threshold=1))
# parallel = FALSE if no parallel backend registered
variants = scoreVarImpactOnGE(variants, epg1, epg2, c('HepG2','K562'))
```

# Example with ChIP-seq

**INCA** can be extended for ChIP-seq data to interrogate transcription factor binding. Here, we provide an example that scores variants using ChIP-seq data for the transcription factor _ATF3_ in [HepG2](https://www.encodeproject.org/experiments/ENCSR000BKE/) and [K562](https://www.encodeproject.org/experiments/ENCSR000BNU/) from [ENCODE](https://www.encodeproject.org/).

## Load required inputs

The required inputs are the same as those needed for RNA-binding proteins, with the exception that shRNA RNA-seq data are unavailable for transcription factors:

+ A variant file with the following fields: _Chr_, _Pos_, _Start_, _End_, _Ref_, _Alt_, and _Gene_.
+ Whole-genome sequencing data of two different cell lines/types with the following fields: _Chr_, _Pos_, _Ref_, _Alt_, _Zyg_.
+ bedNarrow peak files: _Chr_, _Start_, and _End_.

```{r}
directory = 'https://raw.github.com/jduan607/INCA/master'

## GWAS
variants = fread(file.path(directory, 'GWAS', 'final_data.txt'))

## WGS
wgs1 = fread(file.path(directory,'ENCODE_WGS','HepG2_WGS.txt.gz'))
wgs2 = fread(file.path(directory,'ENCODE_WGS','K562_WGS.txt.gz'))

## ENCODE - ATF3 ChIP-seq peaks
peak1 = fread('https://www.encodeproject.org/files/ENCFF725ZKM/@@download/ENCFF725ZKM.bed.gz')[,c(1,2,3,7)]
peak2 = fread('https://www.encodeproject.org/files/ENCFF251XOI/@@download/ENCFF251XOI.bed.gz')[,c(1,2,3,7)]
colnames(peak1) = c('Chr','Start','End','signalValue_IDR')
colnames(peak2) = c('Chr','Start','End','signalValue_IDR')
```

## Align genotypes of K562 and HepG2 cell lines to variants

```{r}
variants = alignCellGenoToVar(variants, wgs1, wgs2, c('HepG2','K562'))
```

## ClinVar-quantiled DeepSEA scores

DeepSEA (Zhou and Troyanskaya, 2015) is a deep learning-based framework trained on genome-wide chromatin profiles, including TF binding, DNase I–hypersensitive sites, and histone marks, to predict the chromatin effects of variant. INCA transforms DeepSEA scores of a variant to represent proportion of pathogenic ClinVar variants that have scores smaller than that of the given variant.

```{r}
ds = fread(file.path(directory,'GWAS','deepsea_results.txt.gz')) # DeepSea scores
ds_quantile = fread(file.path(directory,'ClinVar_scores','clinvar_pathogenic_snv_ds_quantile.txt.gz')) # Quantiles

variants = scoreClinVarQSW(variants, ds, 'ATF3', empirical=ds_quantile)
```

## ChIP-seq allelic effects

We obtained IDR thresholded peak files from ENCODE. For each variant with an allelic difference between the cell lines (HepG2 and K562 in this example), INCA computes a score of allelic effect of TF binding. This score is set to 1 if the variant of interest resides within a binding region of the TF as defined by an IDR peak in one cell line, but not in the other cell line.

```{r}
epg1 = list(list(peaks=peak1))
epg2 = list(list(peaks=peak2))

# parallel = FALSE if no parallel backend registered
variants = scoreAllelicEffect(variants, epg1, epg2, c('HepG2','K562'), window=15, parallel=TRUE)
```
